The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-04 kalman_walk_paper.tex
added citation
edited intro

\section*{Rapid speciation despite conservation of phenotype}
{\centering
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad 
$^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad 
\texttt{plr@uoregon.edu}
\\
}

@@

\begin{abstract}
%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
%the underlying mechanism may have changed,
%since distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small mutational tweaks to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
In this setting we treat an organism as a ``black box''
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
Within this framework, we show that there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct and mutationally connected mechanisms
can lead to the reproductive incompatibility of idependently evolving populations.
  This evolutionary exploration, which we refer to as system drift, proceeds at a rate proportional to the effective population size ($N_e$) and amount of intrapopulation genetic variation.
We estimate that %in reasonably numerous species,
  this process can lead to the formation of new species, perhaps in as few as $0.1 N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally yields, what appears to be, a distinct explanation for Haldane's rule, or why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

@@

A complex molecular machinery translates an organism's genome into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
It is an overarching goal of many biological subdisciplines to attain a general understanding of the function and evolution of this molecular machinery. 
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.

The dynamics of the molecular machinery and its interactions with the environment
can be mathematically described as a dynamical system \citep{jaeger2015comet}.
Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired computational and mathematical models of gene regulatory networks,
\citep[e.g.][]{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
% If we allow the reasonable assumption that the genotype-phenotype map 
% can be represented as a system of differential equations \citep{jaeger2015comet}, 
% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner. 
It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments on the ``black box'' are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The fundamental nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases where,
even if a one-to-one mapping between model and behavior is possible in theory,
  even tiny amounts of noise can make inference problems nonidentifiable in practice \jss{cite?}.

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
%where the optimal phenotype remains constant over evolutionary time,
%so that despite strong stabilizing selection for phenotype,
%neutral drift in the underlying genotype is possible.
Applying results from system theory, we present
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
%which gives concrete expectations for how any such
%system can, in principal, undergo system drift and rewire.
Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
Phenotypically equivalent gene networks are not necessarily compatible with one another, as such system drift may result in the reproductive incompatibility between populations isolated for a sufficiently long period of time, even in the absence of any sort of adaptative, selective, or environmental change. Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.
%Consequentially, two populations isolated for a sufficiently long period of time, 
%will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

\section*{Evolutionary system theory}

Here we construct as abstract model to study biological systems, such as gene regulatory networks. The behavior (i.e. temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
%We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
%within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The $m$ ``inputs'' determined exogenously to the system are denoted $u(t)$,
and the $\ell$ ``outputs'' are denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_i(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$;
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations of the other concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ an $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions,
we can write the phenotype as
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

>>>>

{\centering
{\Large \bf Rapid speciation despite conservation of phenotype} \\
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad 
$^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad 
\texttt{plr@uoregon.edu}
\\
}

@@

\begin{abstract}
%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
%the underlying mechanism may have changed,
%since distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small mutational changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
% Treating an organism as a ``black box''
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
This characterization implies that selectively neutral directions in genotype space should be common
% there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of idependently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate inversely proportional the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

\plr{I removed the $0.1 N_e$ because that's only in the "metapopulation" case,
which is where you've probably already got subpopulations acting like species already.}

@@

It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.

The dynamics of the molecular machinery and its interactions with the environment
can be mathematically described as a dynamical system \citep{jaeger2015comet}.
Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired computational and mathematical models of gene regulatory networks,
\citep[e.g.][]{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
% If we allow the reasonable assumption that the genotype-phenotype map 
% can be represented as a system of differential equations \citep{jaeger2015comet}, 
% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner. 
It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments on the ``black box'' are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The fundamental nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases where,
even if a one-to-one mapping between model and behavior is possible in theory,
even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.

Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
but viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
Below, we first apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
system drift may result in reproductive incompatibility between populations 
isolated for a sufficiently long period of time, 
even in the absence of any sort of adaptative, selective, or environmental change. 
We then use quantitative genetic theory to estimate how quickly 
reproductive incompatibility due to system drift will manifest
\plr{we don't need to summarize the results of the paper in the introduction.  the abstract, and maybe the discussion, but not the introduction.}

 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid} looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.
\plr{missing any important references here?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Evolutionary system theory}

We first describe the class of models we study, such as gene regulatory networks. 
The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
% within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The $m$ ``inputs'' determined exogenously to the system are denoted $u(t)$,
and the $\ell$ ``outputs'' are denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_i(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$;
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations of the other concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ an $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions,
we can write the phenotype as
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.
